Improving the Replicability of Psychological Science Through Pedagogy

### 2016-09-23 paper/paper.Rmd
main text final pass

We also describe our process for conducting replications as part of classroom pedagogy. Although mentorship in experimental methods is an important part of the standard advising relationship, the classroom context allows for elucidation of general principles of good research and discussion of how they can be modified to fit specific instances. And replication research in particular illustrates a number of important concepts -- experimental design, power analysis, reporting standards, and preregistration, among others -- in a way that more open-ended projects, which require new conceptual development, often do not [see @frank2012 for a more extended argument]. There are significant limitations on what can be done in a single term, within the constraints of a course budget and the instructors' expertise. Nevertheless, were this approach implemented more widely, we believe the dividends paid to the field as a whole would be considerable.

@@

A summary of the original experiments chosen for replication is given in Table \ref{tab:summary}. Students chose findings from a wide variety of domains for replication. One cluster of studies included investigations of memory and visual attention [e.g., @xu2015], for example in applied contexts like remembering "deleted" files [@storm2015] or tracking transition between locations on smartphone apps [@liverence2015]. A second cluster came from studies of social perception and judgment, including judgments of faces and voices [@ko2015;@sofer2015], as well as studies of attributions of modesty, creativity, and expertise [@atir2015;@scopelliti2015;@proudfoot2015]. Other studies investigated increasing retirement savings by orienting participants to the future [@lewis2015], legacy priming [@zaval2015], and the effects of math anxiety on performance [@wang2015]. Details of individual replications are available in SOM.

@@

We eventually used a hybrid, case-by-case selection criterion that attempted to maximize the success of the project while staying within the constraints of our budget. For all criteria, experimenters recruited an additional 5% of participants to ensure we still achieved desired power if some participants skipped through or did not complete the experiment.
^[We initially were allocated \$1,500 but one student contributed personal research funds, leading to a total cost of $\approx$ \$1,700 for all studies. We attempted to set payment for our studies at approximately \$6/hour based on timing estimated from Pilot A sessions.] 
In three cases, we powered the replication attempt to 80\% post-hoc power. In four
cases, we used the original sample size. In one case, the original experiment was extremely over-powered in the post-hoc analysis, so we chose a smaller sample size that still would yield high power. In two others, we reduced the number of trials as part of the adaptation to Mechanical Turk (in one of these, changing the sample size to compensate). In the last case, the original study used a multi-step procedure where the first step consisted of stimulus generation; we decreased the N in the first stage of the procedure and focused on achieving high power at the second stage.

@@

All materials and methods for our replication studies are available at [https://github.com/StanfordPsych254](). All experiments were coded in JavaScript, HTML, and CSS so as to be run in a standard web browser. All analyses were written using R, a free and open-source platform for statistical analysis. A schematic of our class workflow is shown in Figure 1.

@@

Students wrote their final reports using [R Markdown](http://rmarkdown.rstudio.com/), a "literate programming" environment in which statistical analysis code can be interspersed with text, figures, and tables. This part of our workflow was extremely useful both pedagogically and for encouraging reproducible research practices. Using the template developed by @osc2015 for replication reports, students created dynamic documents that included their proposed replication study, the code necessary to analyze their data, and the outputs of their data analysis (e.g., statistical tests and figures). This method of writing allows students to share a single compiled document via a hyperlink, facilitating review of writing, results, and code together in a single platform.^[The current manuscript is written in this fashion as well.] This writing method is also likely to reduce the frequency of statistical reporting errors [which appear to be regrettably common; @nuijten2015], given that errors are often introduced by transferring results between statistics and word-processing software packages.

@@

Once requested changes were made, the student conducted "Pilot B," using a handful of naive participants recruited from Mechanical Turk. The goal of this pilot was to ensure that participants were able to complete the experiment and did not report any substantial issues with instructions or technical details of the experiment. (All students were instructed to give participants a way to leave free-form comments at the end of the experiment but prior to debriefing). At the conclusion of Pilot B, both the instructor and a TA reviewed the student's analytic code and its outputs on the data for all confirmatory analyses. All planned analyses were specified on the basis of the key test selected in the original article, and included all exclusion criteria, manipulation checks, etc. specified by the authors. The goal of this code review was to ensure that all planned analyses were specified correctly and with sufficient detail to permit inclusion in the broader analysis.

After Pilot B review was completed, students were given authorization to collect the full sample of participants, contingent on having made any requested changes. Prior to data collection, the analytic script containing all confirmatory analyses was pre-registered using the Open Science Framework. In addition, prior to data collection, we pre-registered the confirmatory analyses reported in the current paper ([https://osf.io/rxz8m/](https://osf.io/rxz8m/)).

@@

Of the 15 students in the class, all completed projects (two with extensions going beyond the class period), one chose a project outside of the sampling frame, and three opted out of the broader project prior to data collection.^[Note that these students did not express doubts about the success of their projects prior to opting out, and 2/3 of these projects were judged by the instructor team to have been successful replications. Thus we have no evidence that these students opted out systematically due to their belief that their project would fail.] The remaining 11 students contributed code, data, registrations and materials to the final product. (In one case, a student did not complete the pre-registration procedure correctly but still submitted a pre-specified analysis plan to the instructors for review. We included this project in the final analysis.)

Each member of the instructor team coded the fidelity of replications, weighing whether the materials, sample, and task parameters differed from the original studies. We coded projects independently on a scale of 1 -- 7 (with 1 being a loose replication with substantive deviations from the original, and 7 being essentially identical), and then discussed our ratings and made adjustments (without coming to full consensus). The mean rating was `r round(mean(d$fidelity), digits = 2)` (Range = `r round(min(d$fidelity), digits = 2)` to `r round(max(d$fidelity), digits = 2)`). Several studies were essentially identical to the originals, but there was a group of others that included differences in population or number of trials that might plausibly have had an effect on the results.

@@

We next turn to an assessment of the replication $p$-value to determine whether the key statistic found a significant effect at the traditional $p=.05$ level. Only `r sum(d$rep_p_value_05)` of the studies yielded a significant replication $p$-value. The well-documented "dance of the p-values" [@cumming2013] suggests that the inference drawn from a single replication p-value may not be informative, however, and other metrics such as effect sizes and Bayes factors may provide a more nuanced perspective.

@@

Effect sizes and Bayes factors improve upon inferences drawn from $p$-values in different ways. The effect size is an estimate of the *magnitude* of an effect, which is stable across sample sizes (assuming a constant experimental procedure, of course) and allows for a more fine-grained comparison across studies than a binary significance criterion. The Bayes Factor is an alternative hypothesis testing method that quantifies the *evidence* in favor of a hypothesis [@jeffreys1961]. Among other advantages, it does not privilege the null hypothesis and can provide evidence either *for* or *against* [see also @scheibehenne2016;@wagenmakers2016].  While the Bayes Factor often agrees with the $p$-value as to which hypothesis is more likely [@wetzels2011], they often substantially disagree on the strength of the evidence.

These additional measures therefore give another sense in which replications can be successful: If we find roughly the same effect size or Bayes factor despite $p$-values on different sides of the arbitrary $p = 0.05$ threshold, we should increase our confidence in the effect's replicability. To compute a standardized effect size, we followed @osc2015 in converting test statistics to a measure of "correlation coefficient per df," which is bounded between 0 and 1^[For example, the conversion from a $t$ statistic is given by $r = \sqrt{t^2/(t^2 + df)}$, where $df$ is the degrees of freedom. Other formulae are given in @osc2015, Supplemental Information.]. We first compared effect sizes from the original and replication studies (Figure 2A). Effect sizes were highly correlated between the two sets of studies 
(r = `r signif(cor.test(d$original_ES_d,d$rep_es)$estimate, digits =2)`, p < .0001), but they were generally smaller in the replications 
(`r signif(coef(lm(d$rep_es~d$original_ES_d-1))[1], digits = 2)*100`\% of the original; 95\% CI [`r signif(confint(lm(d$rep_es~d$original_ES_d-1))[1], digits = 2)*100` - `r signif(confint(lm(d$rep_es~d$original_ES_d-1))[2], digits = 2)*100`]).

While we preregistered an analysis determining whether the original reported effect size is included in the 95\% confidence interval found in the replication, we instead opted for a straightforward comparison of point estimates. This choice was motivated by recent concerns that using confidence intervals in this manner may lead to misleading interpretations [@morey2016], as well as discrepancies in the authors' reporting of effect sizes across the articles we included (which made computing confidence intervals difficult in some cases).

@@

```{r, fig.cap="Replication effect size (A) and Bayes factor (B), plotted by the original effect size and Bayes factor, respectively. Point size shows replication N (A) and ratio of test degrees of freedom (B), color indicates subjective replication assessments by the authors. Note that the key statistic for one replication was a multivariate $F$ test, hence a comparable default Bayes factor could not be computed. Additionally, the original Bayes factor for another was many orders of magnitude greater than the others so it is not displayed.", fig.pos="ht", fig.height=3.6}

@@

In this paper, we reported the results of a series of 11 student replications of previously-published experiments from the 2015 volume of *Psychological Science*. We had two interlocking goals: first, to create a proof of concept for pedagogical replications, and second, to provide evidence on the ability of students replicate important, recent findings in a top journal. Importantly, rather than attempting to assess the truth of particular effects, our project aimed to assess the challenges of replicating findings -- selected on grounds of feasibility and interest -- within the constraints of a course project. Despite this narrower scope, we believe that insights from our approach are applicable to the broader goal of assessing replicability at scale [cf. @osc2015]

@@

From a purely statistical point of view, our results were underwhelming. Despite our relatively close adherence to the original experimental protocols and relatively large sample sizes, the modal outcome of our projects was partial replication. These partial replications often were cases in which some hint of the original pattern was observed but the key statistical test was not statistically significant and showed both smaller effect size and lower evidential value than the same test in the original. We invite readers to browse the narrative descriptions of individual replication attempts in our SOM to see the ways that patterns of experimental findings can differ from one another beyond the significance of a single test.

These statistical findings mask a more optimistic message, however: In most of our projects, the next step for a motivated experimenter is clear. For example, in some projects we suspect that a followup could find strong evidence for the phenomenon of interest by titrating the particular planned analysis or the difficulty of the stimulus materials [@liverence2015;@storm2015;@xu2015]. In others, the difficulty appeared to be statistical power, since differences were in the predicted direction and sometimes reached significance in subsidiary analyses, so a followup would likely require a larger sample [@zaval2015;@lewis2015;@scopelliti2015]. And in some, differences of population would mean that followups would require further stimulus or task adaptation [@sofer2015;@wang2015]. More generally, we believe that our work here underscores the importance of iterated replication [@lewis2016].

Many statistical factors leading to lowered replicability have been discussed in past work, including analytic flexibility, context dependency, publication bias, and low statistical power among others [e.g., @button2013; @vanbavel2016; @ioannidis2005]. For both reasons of design and scale, our study here cannot disentangle these factors empirically. Nevertheless, the decreases in effect size and evidential value we report seem consistent with two explanations. First, we likely saw a "winner's curse," with initial publications tending to over-estimate effects. Second, we also see evidence that specific details of studies vary from context to context; creating successful adaptations can be challenging [@vanbavel2016]. Both of these explain our observations, and we suspect both are present to some degree.

@@

Indeed, by the standards of clinical trials research, nearly every study we replicated in this report would be classified as exploratory research. No protocol was pre-registered, and the general pattern of evidence across studies was often more important to the authors' conclusions than any particular test in any single study.  Almost every study conducted multiple statistical tests, often associated with several statistical models with differing specifications. And in some cases, there was no conventionally-reported and easily calculable effect size measures [e.g., for mixed- and random-effect models, cf. @bakeman2005]. This set of features -- which, in our experience, are endemic to published psychological work, including our own -- makes replication research within the conventional statistical paradigm extremely challenging.

>>>>

We also describe our process for conducting replications as part of classroom pedagogy. Although mentorship in experimental methods is an important part of the standard advising relationship, the classroom context allows for elucidation of general principles of good research and discussion of how they can be modified to fit specific instances. And replication research in particular illustrates a number of important concepts -- experimental design, power analysis, reporting standards, and preregistration, among others -- more directly than open-ended projects, which require new conceptual development [see @frank2012 for extended argument]. There are significant limitations on what can be done in a single term, within the constraints of a course budget and the instructors' expertise. Nevertheless, were this approach implemented more widely, we believe the dividends paid to the field as a whole would be considerable.

@@

A summary of the original experiments chosen for replication is given in Table \ref{tab:summary}. Students chose findings from a wide variety of domains for replication. One cluster of studies included investigations of memory and visual attention [e.g., @xu2015], for example in applied contexts like remembering "deleted" files [@storm2015] or tracking transitions between locations on smartphone apps [@liverence2015]. A second cluster came from studies of social perception and social judgment, including judgments of faces and voices [@ko2015;@sofer2015], as well as studies of attributions of modesty, creativity, and expertise [@atir2015;@scopelliti2015;@proudfoot2015]. Other studies investigated increasing retirement savings by orienting participants to the future [@lewis2015], legacy priming [@zaval2015], and the effects of math anxiety on performance [@wang2015]. Details of individual replications are available in SOM.

@@

We eventually used a hybrid, case-by-case selection criterion that attempted to maximize the success of the project while staying within the constraints of our budget. For all criteria, experimenters recruited an additional 5% of participants to ensure we still achieved the desired criterion if some participants skipped through or did not complete the experiment.
^[We initially were allocated \$1,500 but one student contributed personal research funds, leading to a total cost of $\approx$ \$1,700 for all studies. We attempted to set payment for our studies at approximately \$6/hour based on timing estimated from Pilot A sessions.] 
In four cases, we powered the replication attempt to 80\% post-hoc power. In four
cases, we used the original sample size. In one case, the original experiment was extremely over-powered in the post-hoc analysis, so we chose a smaller sample size that still would yield high power. In two others, we reduced the number of trials as part of the adaptation to Mechanical Turk (in one of these, changing the sample size to compensate). In the last case, the original study used a multi-step procedure where the first step consisted of stimulus generation; we decreased the N in the first stage of the procedure and focused on achieving high power at the second stage.

@@

All experiments were coded in JavaScript, HTML, and CSS so as to be run in a standard web browser. All analyses were written using R, a free and open-source platform for statistical analysis. A schematic of our class workflow is shown in Figure 1.

@@

Students wrote their final reports using [R Markdown](http://rmarkdown.rstudio.com/), a "literate programming" environment in which statistical analysis code can be interspersed with text, figures, and tables. This part of our workflow was extremely useful both pedagogically and for encouraging reproducible research practices. Using the template developed by @osc2015 for replication reports, students created dynamic documents that included their proposed replication study, the code necessary to analyze their data, and the outputs of their data analysis (e.g., statistical tests and figures). This method of writing allows students to share a single compiled document via a hyperlink, facilitating review of writing, results, and code together in a single document.^[The current manuscript is written in this fashion as well.] This writing method is also likely to reduce the frequency of statistical reporting errors [@nuijten2015], given that errors are often introduced by transferring results between statistics and word-processing software packages.

@@

Once requested changes were made, the student conducted "Pilot B," using a handful of naive participants recruited from AMT. The goal of this pilot was to ensure that participants were able to complete the experiment and did not report any substantial issues with instructions or technical details of the experiment. (All students were instructed to give participants a way to leave free-form comments at the end of the experiment, prior to debriefing). At the conclusion of Pilot B, both the instructor and a TA reviewed the student's analytic code and its outputs on the data for all confirmatory analyses. The goal of this code review was to ensure that all planned analyses were specified for the key test selected in the original article, and that all relevant exclusion criteria, manipulation checks, etc. specified by the original authors were implemented correctly.

After Pilot B review was completed, students were given authorization to collect the full sample of participants, contingent on having made any requested changes. Prior to data collection, the analytic script containing all confirmatory analyses was pre-registered using the Open Science Framework. In addition, prior to data collection for all projects, we pre-registered the confirmatory analyses reported in the current paper ([https://osf.io/rxz8m/](https://osf.io/rxz8m/)).

@@

All of the 15 students in the class completed projects (two with extensions going beyond the class period). Of these, one chose a project outside of the sampling frame, and three opted out of the broader project prior to data collection.^[Note that these students did not express doubts about the success of their projects prior to opting out, and 2/3 of these projects were judged by the instructor team to have been successful replications. Thus we have no evidence that these students opted out systematically due to their belief that their project would fail.] The remaining 11 students contributed code, data, registrations, and materials to the final product. (In one case, a student did not complete the pre-registration procedure correctly but still submitted a pre-specified analysis plan to the instructors for review. We included this project in the final analysis.)

Each member of the instructor team coded the fidelity of replications, weighing whether the materials, sample, and task parameters differed from the original studies. We coded projects independently on a scale of 1 -- 7 (with 1 being a loose replication with substantive deviations from the original, and 7 being essentially identical), and then discussed our ratings and made adjustments (without coming to full consensus). The mean rating was `r round(mean(d$fidelity), digits = 2)` (range: `r round(min(d$fidelity), digits = 2)` -- `r round(max(d$fidelity), digits = 2)`). Several studies were essentially identical to the originals, but there was a group of others that included differences in population, number of trials, or method that might plausibly have had an effect on the results.

@@

We next turn to an assessment of the replication $p$-value to determine whether the key statistic found a significant effect at the traditional $p=.05$ level. Only `r sum(d$rep_p_value_05)` of the studies (`r round(mean(d$rep_p_value_05)*100)`%) yielded a significant replication $p$-value. The well-documented "dance of the p-values" [@cumming2013] suggests that the inference drawn from a single replication $p$-value may not be informative, however, and other metrics such as effect sizes and Bayes factors may provide a more nuanced perspective.

@@

Effect sizes and Bayes factors improve upon inferences drawn from $p$-values in different ways. The effect size is an estimate of the *magnitude* of an effect, which is stable across sample sizes (assuming a constant experimental procedure, of course) and allows for a more fine-grained comparison across studies than a binary significance criterion. The Bayes Factor is an alternative hypothesis testing method that quantifies the evidence in favor of a hypothesis [@jeffreys1961]. Among other advantages, it does not privilege the null hypothesis and can provide evidence either for or against [see also @scheibehenne2016;@wagenmakers2016].  While the Bayes Factor often agrees with the $p$-value as to which hypothesis is more likely [@wetzels2011], they often substantially disagree on the strength of the evidence.

These additional measures therefore give another sense in which replications can be successful: If we find roughly the same effect size or Bayes factor despite $p$-values on different sides of the arbitrary $p = 0.05$ threshold, we should increase our confidence in the effect's replicability. To compute a standardized effect size, we followed @osc2015 in converting test statistics to a measure of "correlation coefficient per df," which is bounded between 0 and 1.^[For example, the conversion from a $t$ statistic is given by $r = \sqrt{t^2/(t^2 + df)}$, where $df$ is the degrees of freedom. Other formulae are given in @osc2015, Supplemental Information.] We first compared effect sizes from the original and replication studies (Figure 2A). Effect sizes were highly correlated between the two sets of studies 
(r = `r signif(cor.test(d$original_ES_d,d$rep_es)$estimate, digits =2)`, p < .0001), but they were generally smaller in the replications 
(`r signif(coef(lm(d$rep_es~d$original_ES_d-1))[1], digits = 2)*100`\% of the original; 95\% CI [`r signif(confint(lm(d$rep_es~d$original_ES_d-1))[1], digits = 2)*100` - `r signif(confint(lm(d$rep_es~d$original_ES_d-1))[2], digits = 2)*100`]).

While we preregistered an analysis determining whether the original reported effect size is included in the 95\% confidence interval found in the replication, we instead opted for a straightforward comparison of point estimates. This choice was motivated by the lack of consensus on how to conduct such a test [@anderson2016;@gilbert2016], recent concerns that using confidence intervals in this manner may lead to misleading interpretations [@morey2016], and discrepancies in the authors' reporting of effect sizes across the articles we included (which made computing confidence intervals difficult in some cases).

@@

```{r, fig.cap="Replication effect size (A) and Bayes factor (B), plotted by the original effect size and Bayes factor, respectively. Point size shows replication N (A) and ratio of test degrees of freedom (B), color indicates subjective replication assessments by the authors. Note that the key statistic for one replication was a multivariate $F$ test, hence a comparable default Bayes factor could not be computed. Additionally, the original Bayes factor for oen finding was many orders of magnitude greater than the others so it is not displayed.", fig.pos="ht", fig.height=3.6}

@@

In this paper, we reported a series of 11 student replications of previously-published experiments from the 2015 volume of *Psychological Science*. We had two interlocking goals: first, to create a proof of concept for pedagogical replications, and second, to provide evidence on the ability of students replicate important, recent findings in a top journal. Importantly, rather than attempting to gauge the truth of particular effects, our project aimed to assess the challenges of replicating findings -- selected on grounds of feasibility and interest -- within the constraints of a course project. Despite this narrower scope, we believe that insights from our approach are applicable to the broader goal of assessing replicability at scale.

@@

From a purely statistical point of view, though, our results were underwhelming. Despite our relatively close adherence to the original experimental protocols and relatively large sample sizes, the modal outcome of our projects was partial replication. These partial replications often were cases in which some hint of the original pattern was observed but the key statistical test was not statistically significant and showed both smaller effect size and lower evidential value than the same test in the original. We invite readers to browse the narrative descriptions of individual replication attempts in our SOM to see the ways that patterns of experimental findings can differ from one another beyond the significance of a single test.

These statistical findings mask a more optimistic message, however: In most of our projects, the next step for a motivated experimenter is clear. For example, in some projects we suspect that a followup could find strong evidence for the phenomenon of interest by titrating the particular planned analysis or the difficulty of the stimulus materials. In others, the difficulty appeared to be statistical power, since differences were in the predicted direction and sometimes reached significance in subsidiary analyses, so a followup would likely require a larger sample. And in some, differences of population would mean that followups would require further stimulus or task adaptation. More generally, we believe that our work here underscores the importance of iterated replication [@lewis2016].

Many statistical factors leading to lowered replicability have been discussed in past work, including analytic flexibility, context dependency, publication bias, and low statistical power among others [e.g., @button2013; @vanbavel2016; @ioannidis2005]. For both reasons of design and scale, our study here cannot disentangle these factors empirically. Nevertheless, the decreases in effect size and evidential value we report seem consistent with two explanations. First, we likely saw a "winner's curse," with initial publications tending to over-estimate effects. Second, we also might have seen a reduction in effect size because the original studies were tailored to their specific context and population, whereas our replications may not have been [@vanbavel2016]. Both of these explain our observations, and we suspect both effects are present to some degree.

@@

Indeed, no experiment we attempted to replicate was pre-registered (though some were internal replications), and the general pattern of evidence across studies was often more important to the authors' conclusions than any particular test in any single study.  Almost every study conducted multiple statistical tests, often associated with several statistical models with differing specifications. And in some cases, there was no conventionally-reported and easily calculable effect size measures [e.g., for mixed- and random-effect models, cf. @bakeman2005]. This set of features -- which, in our experience, are endemic to published psychological work, including our own -- makes replication research within the conventional statistical paradigm extremely challenging.
